Inference of a universal social scale and segregation measures using social connectivity kernels

### 2020-08-05 manuscript/main.tex
Add qualifier regarding unobserved dimensions.
Incorporate Sahil's comments.

Building on the successes of conditionally-independent edge models~\cite{Snijders2011} and, in particular, latent space models for social networks~\cite{Hoff2002,Handcock2007,Raftery2012}, we consider a generative model for social networks whose members occupy a multidimensional Blau space in \cref{sec:model}. We discuss desirable properties for social segregation statistics, and, using the generative network model, we develop a suite of statistics applicable to arbitrary attributes in \cref{sec:model-based-segregation}. The statistics capture segregation at different scales: single individuals, pairs of individuals, and society as a whole. Because of both their probabilistic foundations and their construction from the universal notion of a social tie, the segregation statistics have a universal scale, i.e.\ one unit of segregation has the same implications across different societies or at different times. We show that the segregation statistic for pairs of individuals is a semi-metric and can be used to quantify distance in Blau space. We illustrate the statistics with a simple example, and we show that it reduces to a well-known segregation statistic if the attributes are univariate and categorical.

@@

For example, consider a stochastic block model (SBM)~\cite{Snijders2011} with $K$ blocks, intra-group connection probability $\rho_\mathrm{same}$, and inter-group connection probability $\rho_\mathrm{different}<\rho_\mathrm{same}$. Substituting into \cref{eq:social-separation}, the social separation between two nodes with block membership $x$ and $y$ is
\begin{equation}
    \varphi(x, y) = \left(1-\delta_{xy}\right)\left(\logit\rho_\mathrm{same}-\logit\rho_\mathrm{different}\right),\label{eq:social-separation-sbm}
\end{equation}
$\delta_{xy}$ is the Kronecker delta. The social separation only depends on block membership, and it is not affected by the size of each block. For members of different blocks, $\varphi$ is the difference of log odds ratios for the existence of intra-group ties as opposed to inter-group ties previously considered by \textcite{Moody2001}.

@@

To understand how segregated society is as a whole, we would like to aggregate the social isolation $\phi$, but the appropriate statistic depends on the question at hand. For example, if we wanted to study the most isolated subpopulation of society, we should consider $\max_{x\in\blauspace}\phi(x)$. Here, we take a utilitarian approach and, in line with \cref{eq:segregation-individual}, define the \emph{social strain} as
\begin{equation}
    \Phi = \int dx\ P(x) \phi(x),\label{eq:social-strain}
\end{equation}
which quantifies the average social separation amongst members of the society. It is zero when individuals do not discriminate based on attributes, and it can reach arbitrarily large values in a society comprising multiple groups that are completely disconnected. For the SBM, we substitute \cref{eq:segregation-individual-sbm} into \cref{eq:social-strain} and obtain
\begin{align}
    \Phi&= \gamma\left(\logit\rho_\mathrm{same}-\logit\rho_\mathrm{different}\right),\nonumber \\
    \text{where }\gamma&=\sum_{x=1}^K P(x) \left(1 - P(x)\right)\label{eq:segregation-distributional-factor}
\end{align}
is a factor accounting for the relative sizes of the $K$ blocks. The social strain is maximal when the group sizes are equal. If one of the blocks is larger, the strain approaches zero as the sizes of the minority blocks decrease: members of the majority group experiences little social isolation. It is unsurprising that there is no social strain if the society is homogeneous, but the utilitarian approach has a serious limitation: it has little concern for minorities that are not well integrated in society. For equal group sizes, the social strain increases with the number of groups, asymptotically reaching a maximum value of $\logit\rho_\mathrm{same}-\logit\rho_\mathrm{different}$.

@@

In other words, the social separation statistic is a true measure of \emph{distance} in the social space with a probabilistic interpretation if features are themselves measures of distance, including all the features we consider subsequently. This observation puts Peter Blau's~\cite{Blau1977} hypothesis that ``the macrostructure of societies can be defined as a multidimensional space of social positions among which people are distributed and which affect their social relations'' on a sound statistical footing: \emph{fitting conditionally-independent edge models allows us to learn the metric of Blau space}. The metric has a universal scale: \emph{one unit of social separation has the same probabilistic meaning independent of the society under consideration}, facilitating comparison across disparate datasets. Even if two societies have different Blau space dimensions, e.g.\ a society might exist that strongly discriminates based on characteristics which are not found in other societies, the social separation between a pair of individuals has a common meaning.

@@

A representative sample of dyads between individuals together with their demographic attributes is not generally available. However, a number of surveys have collected information about the social ties of respondents using name-generator questions which elicit social ties by asking respondents to nominate their friends~\cite{Kalmijn2007}, individuals they feel close to~\cite{Hipp2009}, or discussion partners~\cite{Marsden1987,McPherson2006}. To generate examples of disconnected dyads, we consider a random sample of pairs of individuals. To account for this non-ignorable data collection process, we introduce a variable $I_{ij}\in\{0,1\}$ indicating whether a particular dyad $A_{ij}$ was observed. The available data thus comprise the demographic attributes $x$ of individuals included in the sample and the dyad state $A_{ij}$ (1 if $i$ and $j$ are connected and $0$ otherwise) if it was observed, i.e.\ $I_{ij}=1$. Adapting the argument presented by \textcite{King2001} to a Bayesian paradigm, we consider the posterior distribution over kernel parameters $\theta$ given the available data:
\begin{equation}
    P(\theta|\observed{A}, \observed{f}, I=1)\propto P(\observed{A}|\theta,\observed{f},I=1)P(\theta),\label{eq:parameter-posterior}
\end{equation}
where $P(\theta)$ is the kernel parameter prior, and $f=f(x,y)$ are features sufficient to evaluate the connectivity kernel given demographic attributes $x$ and $y$. The observed-data likelihood is
\begin{equation}
    P(\observed{A}|\theta,\observed{f},I=1)=\frac{P(\observed{f}|\observed{A},\theta,I=1)P(\observed{A}|\theta,I=1)}{P(\observed{f}|\theta,I=1)}.\label{eq:observed-likelihood}
\end{equation}
Considering the first term in the numerator of \cref{eq:observed-likelihood}, we note that the distribution over kernel features given the state $A$ of the dyad does not depend on whether it was included in the sample or not. More formally,
\begin{align}
    P(f|A,\theta,I=1)&=P(f|A,\theta)\\
    &=\frac{P(A|f,\theta)P(f|\theta)}{P(A|\theta)}.\label{eq:conditional-equivalence}
\end{align}
Turning to the denominator in \cref{eq:observed-likelihood}, we find
\begin{align}
    P(f|\theta,I=1)&=\sum_{\alpha=0}^1 P(f|A=\alpha,\theta,I=1)P(A=\alpha|\theta,I=1)\nonumber\\
    &=P(f|\theta)\sum_{\alpha=0}^1 P(A=\alpha|f,\theta)\frac{P(A=\alpha|\theta,I=1)}{P(A=\alpha|\theta)},\label{eq:likelihood-denominator}
\end{align}
where we used the identity in \cref{eq:conditional-equivalence} to arrive at the second line. Substituting \cref{eq:conditional-equivalence,eq:likelihood-denominator} into \cref{eq:observed-likelihood}, the observed-data likelihood is
\begin{align}
    P(A|\theta,f,I=1)&=\frac{P(A|f,\theta)r(A)}{\sum_{\alpha=0}^1 P(A=\alpha|f,\theta)r(\alpha)},\label{eq:observed-likelihood-final}\\
    \text{where }r(\alpha)&=\frac{P(A=\alpha|\theta,I=1)}{P(A=\alpha|\theta)}\nonumber
\end{align}
is the ratio of prevalences of dyad state $\alpha$ in the sample and the general population. In practice, we approximate the prevalence ratio $r$ using the empirical sample prevalence and prior knowledge about the prevalence in the population. The posterior can be evaluated by substituting \cref{eq:observed-likelihood-final} into \cref{eq:parameter-posterior}, and we can thus infer the parameters $\theta$ from ego network data. See \cref{app:numeric-stability} for details on how to evaluate the observed-data log-likelihood in a numerically stable fashion and \cref{app:inference-validation} for a validation of the inference methodology using synthetic data.

@@

Similar to the social separation in \cref{eq:social-separation-logistic}, the social strain can be broken down into components
\begin{equation}
    \Phi_l = \theta_l \int dx\ dy\ \left(f_l(y, y) - f_l(x, y)\right)\label{eq:segregation-contributions}
\end{equation}
because it is a linear functional of $\varphi$: each component contributes to the social strain in society. For each of the datasets, we evaluate an estimate of the contributions to the social strain
\[
    \hat\Phi_l = \frac{2\theta_l}{\card{\seeds}(\card{\seeds}-1)} \sum_{i < j\in\seeds:i\neq j} \left[f_l(x_j, x_j)-f_l(x_i, x_j)\right],
\]
where the sum is over all distinct pairs of seeds $\seeds$. The contribution $\Phi_l$ quantifies the average social separation due to feature $l$, and it captures the effect of both the connectivity kernel $\rho(x,y)$ and the attribute distribution $P(x)$ in Blau space: neither is sufficient on its own to quantify segregation. Furthermore, $\Phi$ and its contributions in \cref{eq:segregation-contributions} can facilitate comparison across different datasets: they have an intuitive interpretation (the probability of edges decreases with increasing segregation) and universal scale (one unit of segregation has the same effect on edge probability).

@@

>>>>

Building on the successes of conditionally-independent edge models~\cite{Snijders2011} and, in particular, latent space models for social networks~\cite{Hoff2002,Handcock2007,Raftery2012}, we consider a generative model for social networks whose members occupy a multidimensional Blau space in \cref{sec:model}. We discuss desirable properties for social segregation statistics, and, using the generative network model, we develop a suite of statistics applicable to arbitrary attributes in \cref{sec:model-based-segregation}. The statistics capture segregation at different scales: single individuals, pairs of individuals, and society as a whole. Because of both their probabilistic foundations and their construction from the universal notion of a social tie, the segregation statistics have a universal scale, i.e.\ one unit of segregation has the same implications across different societies or at different times. We show that the segregation statistic for pairs of individuals is a semi-metric and can be used to quantify distance in Blau space. We illustrate the statistics with a simple example, and we show that it reduces to a well-known segregation statistic if the attributes are univariate and categorical: the natural logarithm of \citeauthor{Moody2001}'s $\alpha$ index~\cite{Moody2001}.

@@

For example, consider a stochastic block model (SBM)~\cite{Snijders2011} with $K$ blocks, intra-group connection probability $\rho_\mathrm{same}$, and inter-group connection probability $\rho_\mathrm{different}<\rho_\mathrm{same}$. Substituting into \cref{eq:social-separation}, the social separation between two nodes with block membership $x$ and $y$ is
\begin{equation}
    \varphi(x, y) = \left(1-\delta_{xy}\right)\left(\logit\rho_\mathrm{same}-\logit\rho_\mathrm{different}\right),\label{eq:social-separation-sbm}
\end{equation}
$\delta_{xy}$ is the Kronecker delta. The social separation only depends on block membership, and it is not affected by the size of each block. For members of different blocks, $\varphi$ is the difference of log odds ratios for the existence of intra-group ties as opposed to inter-group ties. The social separation is equal to the natural logarithm of the $\alpha$-index proposed by \textcite{Moody2001} for categorical attributes $x$, but $\varphi$ is applicable to arbitrary attributes and connectivity kernels.

@@

To understand how segregated society is as a whole, we would like to aggregate the social isolation $\phi$, but the appropriate statistic depends on the question at hand. For example, if we wanted to study the most isolated subpopulation of society, we should consider $\max_{x\in\blauspace}\phi(x)$. Here, we take a utilitarian approach and, in line with \cref{eq:segregation-individual}, define the \emph{social strain} as
\begin{equation}
    \Phi = \int dx\ P(x) \phi(x),\label{eq:social-strain}
\end{equation}
which quantifies the average social separation amongst members of the society. It is zero when individuals do not discriminate based on attributes, and it can reach arbitrarily large values in a society comprising multiple groups that are completely disconnected. For the SBM, we substitute \cref{eq:segregation-individual-sbm} into \cref{eq:social-strain} and obtain
\begin{align}
    \Phi&= \gamma\left(\logit\rho_\mathrm{same}-\logit\rho_\mathrm{different}\right),\nonumber \\
    \text{where }\gamma&=1 - \sum_{x=1}^K P^2(x) \label{eq:segregation-distributional-factor}
\end{align}
is an index of dispersion~\cite{Moody2001} and accounts for the relative sizes of the $K$ blocks. The social strain is maximal when the groups are of equal size. If one of the blocks is larger, the social strain and index of dispersion approach zero as the sizes of the minority blocks decrease: members of the majority group experiences little social isolation. It is unsurprising that there is no social strain if the society is homogeneous, but the utilitarian approach has a serious limitation: it has little concern for minorities that are not well integrated in society. For equal group sizes, the social strain increases with the number of groups, asymptotically reaching a maximum value of $\logit\rho_\mathrm{same}-\logit\rho_\mathrm{different}$.

@@

In other words, the social separation statistic is a true measure of \emph{distance} in the social space with a probabilistic interpretation if features are themselves measures of distance, including all the features we consider subsequently. This observation puts Peter Blau's~\cite{Blau1977} hypothesis that ``the macrostructure of societies can be defined as a multidimensional space of social positions among which people are distributed and which affect their social relations'' on a sound statistical footing: \emph{fitting conditionally-independent edge models allows us to learn the metric of Blau space}. The metric has a universal scale: \emph{one unit of social separation has the same probabilistic meaning independent of the society under consideration}, facilitating comparison across disparate datasets. Even if two societies have different Blau space dimensions, e.g.\ a society might exist that strongly discriminates based on characteristics which are not found in other societies, the social separation between a pair of individuals has a common meaning. However, it is worth noting that we cannot quantify segregation due to unobserved dimensions of the social space.

@@

A representative sample of dyads between individuals together with their demographic attributes is not generally available. However, a number of surveys have collected information about the social ties of respondents using name-generator questions which elicit social ties by asking respondents to nominate their friends~\cite{Kalmijn2007}, individuals they feel close to~\cite{Hipp2009}, or discussion partners~\cite{Marsden1987,McPherson2006}. To generate examples of disconnected dyads, we consider a random sample of pairs of individuals. To account for this non-ignorable data collection process, we introduce a variable $I_{ij}\in\{0,1\}$ indicating whether a particular dyad $A_{ij}$ was observed. The available data thus comprise the demographic attributes $x$ of individuals included in the sample and the dyad state $A_{ij}$ (1 if $i$ and $j$ are connected and $0$ otherwise) if it was observed, i.e.\ $I_{ij}=1$. Adapting the argument presented by \textcite{King2001} to a Bayesian paradigm, we consider the posterior distribution over kernel parameters $\theta$ given the available data:
\begin{equation}
    P(\theta|\observed{A}, \observed{f}, I=1)\propto P(\observed{A}|\theta,\observed{f},I=1)P(\theta),\label{eq:parameter-posterior}
\end{equation}
where $P(\theta)$ is the kernel parameter prior, and $f=f(x,y)$ are features sufficient to evaluate the connectivity kernel given demographic attributes $x$ and $y$. The observed-data likelihood is
\begin{equation}
    P(\observed{A}|\theta,\observed{f},I=1)=\frac{P(\observed{f}|\observed{A},\theta,I=1)P(\observed{A}|\theta,I=1)}{P(\observed{f}|\theta,I=1)}.\label{eq:observed-likelihood}
\end{equation}
Considering the first term in the numerator of \cref{eq:observed-likelihood}, we note that the distribution over kernel features given the state $A$ of the dyad does not depend on whether it was included in the sample or not. More formally,
\begin{align}
    P(f|A,\theta,I=1)&=P(f|A,\theta)\\
    &=\frac{P(A|f,\theta)P(f|\theta)}{P(A|\theta)}.\label{eq:conditional-equivalence}
\end{align}
Turning to the denominator in \cref{eq:observed-likelihood}, we find
\begin{align}
    P(f|\theta,I=1)&=\sum_{\alpha=0}^1 P(f|A=\alpha,\theta,I=1)P(A=\alpha|\theta,I=1)\nonumber\\
    &=P(f|\theta)\sum_{\alpha=0}^1 P(A=\alpha|f,\theta)\frac{P(A=\alpha|\theta,I=1)}{P(A=\alpha|\theta)},\label{eq:likelihood-denominator}
\end{align}
where we used the identity in \cref{eq:conditional-equivalence} to arrive at the second line. Substituting \cref{eq:conditional-equivalence,eq:likelihood-denominator} into \cref{eq:observed-likelihood}, the observed-data likelihood is
\begin{align}
    P(A|\theta,f,I=1)&=\frac{P(A|f,\theta)r(A)}{\sum_{\alpha=0}^1 P(A=\alpha|f,\theta)r(\alpha)},\label{eq:observed-likelihood-final}\\
    \text{where }r(\alpha)&=\frac{P(A=\alpha|\theta,I=1)}{P(A=\alpha|\theta)}\nonumber
\end{align}
is the ratio of prevalences of dyad state $\alpha$ in the sample and the general population. In practice, we approximate the prevalence ratio $r$ using the empirical sample prevalence and prior knowledge about the prevalence in the population. The posterior can be evaluated by substituting \cref{eq:observed-likelihood-final} into \cref{eq:parameter-posterior}, and we can thus infer the parameters $\theta$ from ego network data. See \cref{app:numeric-stability} for details on how to evaluate the observed-data log-likelihood in a numerically stable fashion and \cref{app:inference-validation} for a validation of the inference methodology using synthetic data. For logistic connectivity kernels, the observed-data likelihood in \cref{eq:observed-likelihood-final} resembles a conventional case-control likelihood, e.g.\ as used by \textcite{Smith2014}.

@@

Similar to the social separation in \cref{eq:social-separation-logistic}, the social strain can be broken down into components
\begin{equation}
    \Phi_l = \theta_l \int dx\ dy\ \left(f_l(y, y) - f_l(x, y)\right)\label{eq:segregation-contributions}
\end{equation}
because it is a linear functional of $\varphi$: each component contributes to the social strain in society. For each of the datasets, we evaluate an estimate of the contributions to the social strain
\[
    \hat\Phi_l = \frac{2\theta_l}{\card{\seeds}(\card{\seeds}-1)} \sum_{i < j\in\seeds:i\neq j} \left[f_l(x_j, x_j)-f_l(x_i, x_j)\right],
\]
where the sum is over all distinct pairs of seeds $\seeds$. The contribution $\Phi_l$ quantifies the average social separation due to feature $l$, and it captures the effect of both the connectivity kernel $\rho(x,y)$ and the attribute distribution $P(x)$ in Blau space: neither is sufficient on its own to quantify segregation. Furthermore, $\Phi$ and its contributions in \cref{eq:segregation-contributions} can facilitate comparison across different datasets, as illustrated in \cref{fig:segregation-embedding}: they have an intuitive interpretation (the probability of edges decreases with increasing segregation) and universal scale (one unit of segregation has the same effect on edge probability).

@@

\section*{Acknowledgements}

We would like to thank Sahil Loomba for useful feedback on the manuscript.
